---
title: "Bank Churners - Analysis of Credit Card customer Attrition"
author: "Archana Nagaraja"
date: "03/02/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is the final project for METCS544 - Bank Churners. 

**Goal of the project:** Analyze the customer attrition data and gather insights about what could be the contributing factors.

Link to dataset : [Bank-Churners-Dataset](https://www.kaggle.com/sakshigoyal7/credit-card-customers)

The dataset contains 8500 records of existing customers and 1627 records of attrited customers.

We will be analyzing the following :-

1. Attrition by Card-Category
2. Months-on-book analysis for attrited customers
3. Income-category/Card-category/Gender analysis for Attrited customers
4. Distribution of customer's Age
5. Review of Central Limit Theorm - Customer's Age
6. Sampling methods - Simple Random, Systematic & Stratified

### Loading the csv file with raw data
```{r}

library(modeest)
options(digits=3)
library(sampling)
library(plotly)

setwd("C:/ARCHANA/Boston University MS Applied Data Analytics/METCS544 - Foundations of Analytics with R - Spring1/Final Project/")

raw.data <- read.csv("BankChurners.csv")
```

### Data Pre-processing

**Retaining only the columns that are being used for analysis**

```{r}

raw.data <- raw.data[,names(raw.data) %in% c("Attrition_Flag","Customer_Age",
                        "Gender","Income_Category",
                        "Card_Category","Months_on_book")]

```

**Looking for missing values in the data**

```{r}
unique(raw.data$Customer_Age)
unique(raw.data$Gender)
unique(raw.data$Income_Category) # has 1112 unknown values
nrow(raw.data[raw.data$Income_Category=="Unknown",])
unique(raw.data$Card_Category)
unique(raw.data$Months_on_book)


```
Income-Category field has unknown values, removing these rows from the raw data

```{r}
raw.data <- subset(raw.data,Income_Category!="Unknown")
```

**Split the raw data into 2 parts - Attrited and Existing Customers**

```{r}
existing.cust <- raw.data[raw.data$Attrition_Flag=="Existing Customer",names(raw.data) %in%
                        c("Attrition_Flag","Customer_Age","Gender","Income_Category",
                          "Card_Category","Months_on_book")]

attrited.cust <- raw.data[raw.data$Attrition_Flag=="Attrited Customer",names(raw.data) %in%
                      c("Attrition_Flag","Customer_Age","Gender","Income_Category",
                        "Card_Category","Months_on_book")]
```
***
### PART-1 - Analysis for categorical variable - card_category

**Card Category** refers to the different type of credit cards that are offered to customers - **Blue, Silver, Gold & Platinum**.

Here is the table representation to view the frequencies of card-category

All Customers -

```{r}

all.card <- table(raw.data$Card_Category)
all.card
```

Attrited customers - 

```{r}

attr.card <- table(attrited.cust$Card_Category)
attr.card

sprintf("Percentage of customers who attrited - by Card Category - %s - %.2f%%",
        rownames(attr.card),round((attr.card/all.card)*100,2))
```

```{r}

card.types <- c("Blue","Gold","Platinum","Silver")

plot_ly() %>%
add_trace(x = ~card.types, y = ~all.card, type = 'bar',
            text = all.card, textposition = 'auto',name="All Customers",
            marker = list(color = 'blue',
            line = list(color = 'blue', width = 1.5))) %>%
add_trace(x = ~card.types, y = ~attr.card, type = 'bar',
            text = attr.card, textposition = 'auto',name="Attrited Customers",
            marker = list(color = 'red',
            line = list(color = 'red', width = 1.5))) %>%
layout(title = "Distribution of Card Categories",
                      barmode = 'group',
                      xaxis = list(title = "Card Category"),
                      yaxis = list(title = "No. of Customers"))


```


**INFERENCES - **

- We see that majority of the attrited(1343) customers held a blue category credit card.

- Among the 4 card categories, Platinum has highest attrition rate - 20%

Analysis of distribution of card categories using barplot

***
### PART-2 - Analysis for numeric variable - months-on-book for attrited customers

**Months-on-book** refers to the number of months the customer has held the credit-card.

Here is a boxplot of months-on-book to analyze the spread

```{r}

f <- fivenum(attrited.cust$Months_on_book)
out <- c(f[2]-1.5*(f[4]-f[2]),f[4]+1.5*(f[4]-f[2]))

plot_ly(attrited.cust, y = ~Months_on_book, type="box", 
        name = 'Months-on-book',quartilemethod="exclusive") %>%
    layout(title="Months-on-book - Attrited Customers")

sprintf("The average months-on-book for attrited customers is %g",
        round(mean(attrited.cust$Months_on_book)))

sprintf("Most of the attrited customers were with the company for %s months",
        mfv(attrited.cust$Months_on_book))

```

**INFERENCES - **

- The months-on-book for attrited customers ranges between 13 and 56 with about 11 outliers.

- The interquartile range/middle 50% of the data is between 32 and 40 and the median of 36 separates top 50% of the data from bottom 50%.

- The average is about the same as the median of 36.

Histogram to analyze frequencies of the months that customers stayed with the company - 


```{r}

hist(attrited.cust$Months_on_book,main="Months-on-book - Attrited 
     Customers",col="light blue",xlab="months-on-book")

```

**INFERENCE - **

- About 600(41%) of the attrited customers were with the company for 35-40 months.

***
### PART-3 - Analysis for set of 2 or more variables - Customer income-category/card-category

**Income Category** refers to the income range of the customer - 

Less than $40K, 
$40K-$60K, 
$60K-$80K, 
$80K-$120K, 
$120K+ 

Contingency table of income-category & card-category - Attrited customers

```{r}

sorted.attr.data <- attrited.cust[order(attrited.cust$Income_Category),]

sorted.attr.data$Income_Category <- factor(sorted.attr.data$Income_Category,
                                          levels=c("Less than $40K","$40K - $60K",
                                          "$60K - $80K","$80K - $120K",
                                          "$120K +"))

income.card <- table(sorted.attr.data$Income_Category,
                     sorted.attr.data$Card_Category)

income.card

```

**INFERENCES - **

- We see that for customers with income of 120k+, none of them were holding a platinum card and very few held Gold cards. Platinum/Gold cards with better rewards/perks could have given these customers more interest to stay.

- For all income-levels, we see a big disparity between number of customers holding blue vs silver - we could look into opportunities for upgrading/reduce attrition.


Marginal & conditional distribution of income-category & card-category

```{r}

addmargins(income.card)

# income-category
round(prop.table(income.card,1),3)

# card-category
round(prop.table(income.card,2),3)

```

**INFERENCE - **

- The total number of platinum/gold customers is much lesser compared to silver. We should look into offering upgrades to motivate customers to stay.

Mosaic plot - representation of contingency table

```{r}

mosaicplot(income.card,color=c("pink","purple"),cex.axis=0.6,las=1,
           xlab="Income Category",ylab="Card Category",
           main="IncomeCategory vs CardCategory")

```


**Analysis - Customer Income-Category/Card-Category/Gender**

Breakdown for attrited customers - 

```{r}
aincome.card.gender <- table(sorted.attr.data$Income_Category,
                            sorted.attr.data$Card_Category,
                            sorted.attr.data$Gender)

ftable(aincome.card.gender)

```

Breakdown for existing customers - 

```{r}

sorted.exis.data <- existing.cust[order(existing.cust$Income_Category),]

sorted.exis.data$Income_Category <- factor(sorted.exis.data$Income_Category,
                                           levels=c("Less than $40K","$40K - $60K",
                                                    "$60K - $80K","$80K - $120K",
                                                    "$120K +"))

eincome.card.gender <- table(sorted.exis.data$Income_Category,
                             sorted.exis.data$Card_Category,
                             sorted.exis.data$Gender)

ftable(eincome.card.gender)

```

Bar Plots for existing vs attrited - by Gender - 


```{r}

male <- raw.data[raw.data$Gender=="M",]
female <- raw.data[raw.data$Gender=="F",]

mpercent <- paste0(round(table(male$Attrition_Flag)/nrow(male),2)*100,"%")
fpercent <- paste0(round(table(female$Attrition_Flag)/nrow(female),2)*100,"%")

subplot(
plot_ly(x=c("Attrited","Existing"),y=table(male$Attrition_Flag),type="bar",
        name="Male") %>%
    add_text(text=mpercent,textposition = "top",showlegend=FALSE),
plot_ly(x=c("Attrited","Existing"),y=table(female$Attrition_Flag),type="bar",
        name="Female") %>%
    add_text(text=fpercent,textposition = "top",showlegend=FALSE) %>%
layout(title="Attrition by Gender"),
shareY = TRUE)


```

***
### PART-4 - Distribution of numeric data - customer's age

Boxplot of customer's age to analyze the spread - 

```{r}

subplot(
plot_ly(existing.cust, y = ~Customer_Age, type="box", 
            name = 'Existing',quartilemethod="exclusive"),
plot_ly(attrited.cust, y = ~Customer_Age, type="box", 
        name = 'Attrited',quartilemethod="exclusive") %>%
    layout(title="Age Distribution - Existing vs Attrited Customers"),
shareY = TRUE)

sprintf("The average age for attrited customers is %g",
        round(mean(attrited.cust$Customer_Age)))

sprintf("The average age for existing customers is %g",
        round(mean(existing.cust$Customer_Age)))


```

**INFERENCE - **

- The age distribution of existing and attrited customers is approximately the same with few outliers for existing customers.

- The average age for both attrited and existing customers is around 46 which is same as the median.

- This may suggest that age may not be a big contributor to attrition.

Histogram to visualize the distribution - 

```{r}
par(mfrow=c(1,2))

hist(attrited.cust$Customer_Age,main="Attrited Customers",xlab="Customer's Age",
     xlim=c(20,70),col="red")

hist(existing.cust$Customer_Age,main="Existing Customers",xlab="Customer's Age",
     xlim=c(20,80),col="blue")

```

**INFERENCE - **

- The age distribution for both attrited & existing follows a normal distribution.

Probability Density Function - Attrited Customers - 

```{r}

mean.a <- mean(attrited.cust$Customer_Age)

sd.a <- sd(attrited.cust$Customer_Age)

data.a <- dnorm(attrited.cust$Customer_Age,mean=mean.a,sd=sd.a)

x.a <- seq(min(attrited.cust$Customer_Age),max(attrited.cust$Customer_Age),5)

plot(attrited.cust$Customer_Age,data.a,pch=19,main="PDF - Attrited Customers",
     xlab="Customer's Age",ylab="Probability Density Function")

```

Cumulative Density Function - Attrited Customers - 

```{r}
data.b <- pnorm(attrited.cust$Customer_Age,mean=mean.a,sd=sd.a)

plot(attrited.cust$Customer_Age,data.b,pch=19,main="CDF - Attrited Customers",
     xlab="Customer's Age",ylab="Cumulative Density Function")

```

***
### PART-5 - Central Limit Theorm - Customer's Age

We have taken 5000 samples of sizes 10,20,30 & 40 and computed sample-means for the same.

```{r}

set.seed(2633)

data.mean <- round(mean(raw.data$Customer_Age),2)
data.sd <- round(sd(raw.data$Customer_Age),2)

# function to pick 5000 samples and compute sample-means
sample.func <- function(sample.size,sample.count) {
    sample.means <- numeric(sample.count)
    i <- 1
    for (i in 1:sample.count) {
        sample.means[i] <- mean(sample(raw.data$Customer_Age,sample.size,
                                       replace = FALSE))
    }
    return (sample.means)
}

# sample-size: 10
sample.means10 <- sample.func(10,5000)
mean.sm10 <- mean(sample.means10)
sd.sm10 <- sd(sample.means10)

# sample-size: 20
sample.means20 <- sample.func(20,5000)
mean.sm20 <- mean(sample.means20)
sd.sm20 <- sd(sample.means20)

# sample-size: 30
sample.means30 <- sample.func(30,5000)
mean.sm30 <- mean(sample.means30)
sd.sm30 <- sd(sample.means30)

# sample-size: 40
sample.means40 <- sample.func(40,5000)
mean.sm40 <- mean(sample.means40)
sd.sm40 <- sd(sample.means40)
```

We then plot the distribution of sample-means for sample-size: 40

```{r}
# Plot
sd3 <- c(mean.sm40-3*sd.sm40,mean.sm40+3*sd.sm40)
# Plot density of sample means
hist(sample.means40,main="Sample-means - Sample-size: 40",
                               col="blue",prob=TRUE,
                            xlab="Sample means - Customer's Age")
abline(v=sd3,col="red") # 3-sd from the mean

```

**INFERENCE - **

- From the plot, we see that the distribution of sample means follows a normal distribution.  - 
- Most of the data is within 3 standard deviations of the mean.

Analyzing the mean & standard-deviation of the sample-means - 

```{r}

sprintf("Data Mean: %.2f; Data Standard Deviation: %.2f",data.mean,data.sd)

sprintf("Sample-size: %g, Mean: %.2f, SD: %.2f",c(10,20,30,40),
        c(mean.sm10,mean.sm20,mean.sm30,mean.sm40),
        c(sd.sm10,sd.sm20,sd.sm30,sd.sm40))

sprintf("Data SD/sqrt(sample.size): %.2f; SD-SampleMeans: %.2f",
        (data.sd/sqrt(40)),sd.sm40)

```

**INFERENCES - **

- Mean of data is about equal to the mean of the sample-means for all sample-sizes.

- Standard Deviation decreases as the sample-size increases.

- Standard Deviation of data/sqrt(sample.size) is equal to Standard Deviation of sample-means.

***
### PART-6 - Sampling Methods

For each of the sampling methods used, we analyze the following - 

1) How reflective is the sample of the actual data - split of existing vs attrited customers, split of card-categories.

2) Which card-category has the highest attrition compared to findings from the actual data.

3) Months-on-book analysis for attrited customers - does the sample reflect the findings from actual data.

**Simple Random Sampling**

Sample-size: 100
```{r}

set.seed(2633)

srs <- srswor(n=100,N=nrow(raw.data))
srs.sample <- raw.data[srs!=0,]

```
1) Reviewing the fraction of existing vs attrited customers picked by the sample - 

```{r}
paste("Fraction of Attrited vs Existing customers in the data - ")
table(raw.data$Attrition_Flag)/nrow(raw.data)

paste("Fraction of Attrited vs Existing customers picked by SimpleRandomSampling")
table(srs.sample$Attrition_Flag)/nrow(srs.sample)
```

Split of card-categories in the sample selected - 

```{r}
table(srs.sample$Card_Category)

```

**INFERENCES - **

-We see that the samples picked through SRSWOR approximately reflects the split between attrited/existing customers.

-However, the sample only contains data for Blue and Silver card categories.

2) Attrition by card-category - 

```{r}
attr.srs <- srs.sample[srs.sample$Attrition_Flag=="Attrited Customer",]

table(attr.srs$Card_Category)
```

All 14 samples picked are under the Blue category.Unable to validate the card-category with highest attrition.

3) Months-on-book for Attrited Customers - 

```{r}
hist(attr.srs$Months_on_book,main="Simple Random Sampling",xlab="Months-on-book",labels=TRUE,col="orange")
```

6/14 = 42% of the attrited customers were with the company - 35-40 months - similar to the findings from the actual data.

**Systematic Sampling**

Sample-size: 100

```{r}
set.seed(2633)

grp <- ceiling(nrow(raw.data)/100) # divide into groups
r <- sample(grp,1) # pick first sample
sys.sample <- raw.data[seq(r, by=grp, length=100),]

```

1) Reviewing the fraction of existing vs attrited customers picked by the sample -
```{r}
paste("Fraction of Attrited vs Existing customers picked by Systematic Sampling")
table(sys.sample$Attrition_Flag)/nrow(sys.sample)
```

We see that the samples picked through Systematic sampling approximately reflects the split between attrited/existing customers, but SRS yielded a closer split.

Split of card-categories in the sample selected - 

```{r}
table(sys.sample$Card_Category)
```

2) Attrition by card-category - 

```{r}
attr.sys <- sys.sample[sys.sample$Attrition_Flag=="Attrited Customer",]

table(attr.sys$Card_Category)
```

3) Months-on-book for Attrited Customers - 

```{r}
hist(attr.sys$Months_on_book,main="Systematic Sampling",xlab="Months-on-book",labels=TRUE,col="purple")
```

7/12=58% of attrited customers were with the company for 35-40 months as per samples picked from systematic sampling.

**Stratified sampling - using proportional sizes based on the Card-category**

```{r}
set.seed(2633)

sorted.data <- raw.data[order(raw.data$Card_Category),]

size <- round(100*table(raw.data$Card_Category)/nrow(raw.data))

paste(c("Size of samples proportional based on card-category - ",size))

# Error Encountered in Strata : Error in data.frame(..., check.names = FALSE) : 
 # arguments imply differing number of rows: 0, 1
# This is due to one of the groups(platinum) having a size=0
# Overriding the size to replace 0 with 1 for platinum
size <- c(92,1,1,6) 

strat <- strata(sorted.data,stratanames="Card_Category",size=size,
                method="srswor")

strata.sample <- getdata(sorted.data,strat)
```

1) Reviewing the fraction of existing vs attrited customers picked by the sample

```{r}
paste("Fraction of Attrited vs Existing customers picked by Stratified Sampling")
table(strata.sample$Attrition_Flag)/nrow(strata.sample)
```

**INFERENCE - **

- The samples picked through Stratified sampling approximately reflects the split between attrited/existing customers in the actual data,but SRS had a closer split.

Split of card-categories in the sample selected - 

```{r}
table(strata.sample$Card_Category)

```

2) Attrition by card-category -

```{r}
attr.str <- strata.sample[strata.sample$Attrition_Flag=="Attrited Customer",]

table(attr.str$Card_Category)

```

The sample does not contain data for platinum/gold attrited customers.

3) Months-on-book for Attrited Customers - 

```{r}
hist(attr.str$Months_on_book,main="Stratified Sampling",xlab="Months-on-book",labels=TRUE,col="pink")
```

4/13=46% of attrited customers were with the company for 35-40 months as per samples picked from systematic sampling.

***
### Conclusion

- Attrition across the 4 card categories closely reflected the customer split across all customers. We can infer that attrition is across the card-categories and will need to be addressed.

- About 41% of the customers who were with the company 35-40 months left the company.

- We found that across all the income categories, majority of the customers owned a Blue card and very few customers owned a Silver/Gold card. This may be contributing to the attrition.

***
### Next Steps

- I would like to explore various sampling methods like Cluster Sampling and Stratified sampling using Systematic method.

- I would also like to analyze details like customer's education-level, marital-status and credit-limit to gather insights about how they may be contributing to attrition.

***
### Thank you